\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2025-05-09 scratch/JSS-submission/Final_submission/jss5535.tex
edit .Rmd, .tex, and update .pdf files

Validating research by replication on independently collected data is a
common scientific norm. Emulating this process in a single study by
data-division is less common \citep[see, e.g.,][]{Barnard:1974}: The
data are randomly divided into two, possibly equal-size, parts; the
first part is used to develop and fit (in the language of machine
leaning, ``train'') a statistical model; and then the second part is
used to assess the adequacy of (``validate'') the model fit to the first
part of the data. Data-division, however, suffers from two problems: (1)
Dividing the data decreases the sample size and thus increases sampling
error; and (2), even more disconcertingly, the results can vary
substantially based on the random division of the data, particularly in
smaller samples. See \citet[Sec. 5.3]{Harrell:2015} for cogent remarks
about data-division and cross-validation.

@@

The \code{cv()} function is introduced in the context of a preliminary
example in Section \ref{preliminary-example-polynomial-regression} of
the paper.

\begin{itemize}
\item
  The \code{"lm"} and \code{"glm"} methods are for linear and
  generalized-linear models fit respectively by the standard
  \proglang{R} \code{lm()} and \code{glm()} functions.
\item
  The \code{"modList"} method for \code{cv()} cross-validates several
  competing models, not necessarily of the same class, using the same
  division of the data into folds.
\item
  Cross-validating mixed-effects models, using the \code{"glmmTMB"},
  \code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves
  special considerations that we take up in Section
  \ref{cross-validating-mixed-effects-models}.
\item
  The \code{"function"} method for \code{cv()}, discussed in Section
  \ref{cross-validating-model-specification}, cross-validates a
  potentially complex model-specification process that may, for example,
  involve choice of data transformations and predictors, or model
  selection via CV itself.
\item
  The \code{"default"} \code{cv()} method works (perhaps with a bit of
  coaxing) with many other existing regression-model classes for which
  there is an \code{update()} method that accepts a \code{data}
  argument. More generally, the \pkg{cv} package is designed to be
  extensible, as discussed in Section \ref{extending-the-cv-package}.
\end{itemize}

@@

{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-2} }

@@

We'll focus on the relationship of \code{mpg} (miles per gallon) to
\code{horsepower}, as displayed in
Figure~\ref{fig:mpg-horsepower-polynomials} (a). The relationship
between the two variables is monotone, decreasing, and nonlinear.
Following \citet{JamesEtAl:2021}, we'll consider approximating the
relationship by a polynomial regression, with the degree of the
polynomial \(p\) ranging from 1 (a linear regression) to 10.\footnote{Although
  it serves to illustrate the use of CV, a polynomial is not the best
  choice here. Consider, for example, the scatterplot for
  log-transformed \code{mpg} and \code{horsepower}, produced by
  \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of
  which is left to the reader). We revisit the \code{Auto} data in
  Section \ref{cross-validating-model-specification}.} Polynomial fits
for \(p = 1\) to \(5\) are shown in Figure
\ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly
inappropriate; the fits for \(p = 2\) (quadratic) through \(4\) are very
similar; and the degree-5 polynomial may over-fit the data by chasing
one or two relatively high \code{mpg} values at the right (but see the
CV results reported below).

Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of
estimated (squared) error as a function of polynomial-regression degree:
The mean-squared error (``MSE''), defined as
\(\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\), and
the usual residual variance, defined as
\(\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2\).
The former necessarily declines with \(p\) (or, more strictly, can't
increase with \(p\)), while the latter gets slightly larger for the
largest values of \(p\), with the ``best'' value, by a small margin, for
\(p = 7\).

@@

{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-1} }~~~\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-2} }

@@

Following their use by \citet{RaudenbushBryk:2002}, data from the 1982
\emph{High School and Beyond} (``HSB'') survey have become a staple of
the literature on mixed-effects models. The HSB data are used by
\citet[Sec. 7.2.2]{FoxWeisberg:2019} to illustrate the application of
linear mixed models to hierarchical data, and we'll closely follow their
example here.

@@

In the process, we merge variables from the school-level and
student-level data sets, and create two new school-level variables:
\code{mean.ses}, which is the average SES for students in each school;
and \code{cses}, which is the individual students' SES centered at their
school means. For details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.

@@

To generate the four data sets, we first constructed a template data set
from a population with a between-patient effect of dosage of \(1.0\) and
a within-patient effect of \(-0.5\). We then applied four different
multipliers to the between-patient standard deviation from the
regression line of patient centroids. The resulting data sets, shown in
Figure \ref{fig:plot1}, differ in the between-patient variation of
patient centroids from a common between-patient regression line,
exhibiting a progression from low to high variation around the common
regression line. Each patient has the same relative configuration of
dosages and symptoms in each of the four data sets. To help visualize
the structure of the data, estimated 50\% Gaussian concentration
ellipses are shown for each patient.

@@

{\centering \includegraphics[width=0.75\linewidth]{Figures/plot1-1}

@@

Using data like these, a researcher may attempt to obtain an estimate of
the within-patient effect of dosage by fitting mixed models with a
random intercept. We will illustrate how this approach results in
estimates that are highly sensitive to the relative variation at the two
levels of the mixed model by considering four models, all of which
include a random intercept. The first three models are sequentially
nested in their fixed effects: (1) \code{~ 1}, intercept only; (2)
\code{~ 1 + x}, intercept and effect of \code{x}; and (3)
\code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual
variable, \code{xm}, consisting of the within-patient mean of \code{x}.
The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the
centered-within-group variable \code{x - xm}. We thus fit four models to
four data sets for a total of 16 models.

We proceed to obtain predictions from each model (1) using only the
fixed effects, as for cross-validation based on clusters (i.e.,
patients), and (2) using both fixed and random effects---that is, the
BLUPs---as for cross-validation based on cases (i.e., occasions within
patients). The fixed- and random-effects predictions from these models
are displayed in Figure \ref{fig:plot-fits-combined} along with a
summary of the data using 50\% concentration ellipses (see the top row
of the figure).

@@

{\centering \includegraphics[width=1\linewidth]{Figures/plot-fits-combined-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{Figures/cross-validation-data-plot-1}

@@

In summary, when between-cluster variation is relatively large, the
model \code{~ 1 + x}, with \code{x} alone and without the contextual
mean \code{xm} of \code{x}, is assessed as fitting very poorly by
cluster-based CV, but relatively much better by case-based CV. In all of
our examples, the model \code{~ 1 + x + xm}, which includes both
\code{x} and its contextual mean, produces better results using both
cluster-based and case-based CV. These conclusions are consistent with
our observations based on graphing predictions from the various models
(in Figure \ref{fig:plot-fits-combined} on page
\pageref{fig:plot-fits-combined}), and they illustrate the desirability
of assessing mixed-effect models at different hierarchical levels.

@@

As \citet[Sec. 7.10.2: ``The Wrong and Right Way to Do
Cross-validation'']{HastieTibshiraniFriedman:2009} explain, if the whole
data are used to specify or fine-tune a statistical model, then
subsequent cross-validation of the model is intrinsically misleading,
because the model is selected to fit the whole data, including the part
of the data that remains when each fold is removed. Statistical modeling
is partly a craft, and one could imagine applying that craft
independently to successive partial data sets, each with a fold removed.
The resulting procedure would be tedious, though possibly worth the
effort, but it would also be difficult to realize in practice: After
all, we can hardly erase our memory of statistical modeling choices
between analyzing partial data sets. Alternatively, if we're able to
automate the process of model specification, then we can more
realistically apply CV mechanically.

@@

\subsection{Example: Data transformation and predictor selection for the
Auto
data}\label{example-data-transformation-and-predictor-selection-for-the-auto-data}

@@

The \code{Auto} data appeared in a preliminary example in Section
\ref{preliminary-example-polynomial-regression}, where we employed CV to
inform the selection of the degree of a polynomial regression of
\code{mpg} on \code{horsepower}. Here, we consider more generally the
problem of predicting \code{mpg} from the other variables in the
\code{Auto} data. We begin with a bit of data management, and then
examine the pairwise relationships among the numeric variables in the
data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by
the \code{scatterplotMatrix()} function in the \pkg{car} package); in
each panel, the solid line shows the linear least-squares fit and the
broken line is for a nonparametric regression:

@@

{\centering \includegraphics[width=0.95\linewidth]{Figures/Auto-explore-scatterplot-matrix-1}

@@

{\centering \includegraphics[width=0.7\linewidth]{Figures/Auto-explore-cr-plots-1}

@@

Some background: As \citet[Sec. 8.2]{Weisberg:2014} explains, there are
technical advantages to having (numeric) predictors in linear regression
analysis that are themselves linearly related. If the predictors
\emph{aren't} linearly related, then the relationships between them can
often be straightened by power transformations. Transformations can be
selected after graphical examination of the data, or by analytic
methods, such as transforming the predictors towards multivariate
normality, which implies linearity. Once the relationships between the
predictors are linearized, it can be advantageous similarly to transform
the conditional distribution of the response variable towards normality.
Selecting transformations analytically raises the possibility of
automating the process, as required for cross-validation.

The \code{powerTransform()} function in the \pkg{car} package transforms
variables towards multivariate normality by a generalization of Box and
Cox's maximum-likelihood-like approach \citep{BoxCox:1964}. Although the
multinormal likelihood is maximized, the variables are transformed
individually. Several ``families'' of power transformations can be used,
including the original Box-Cox family (which is the default), simple
powers (and roots), and two adaptations of the Box-Cox family to data
that may include negative values and zeros: the Box-Cox-with-negatives
family and the Yeo-Johnson family; see \citet[Chap. 8]{Weisberg:2014}
and \citet[Chap. 3]{FoxWeisberg:2019} for details. We proceed to
transform the numeric predictors in the \code{Auto} regression towards
multivariate normality:

@@

\subsection[Example: Applying meta CV to polynomial regression for the
Auto data]{\texorpdfstring{Example: Applying meta CV to polynomial
regression for the Auto
data\footnote{What we call here ``meta CV'' has also been termed
  ``nested CV'' \citep[e.g., in the \pkg{cvms} package
  by][]{OlsenZachariae:2024}. We prefer, however, to reserve the term
  ``nested CV'' for the procedure described by
  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying meta CV to polynomial regression for the Auto data}}\label{example-applying-meta-cv-to-polynomial-regression-for-the-auto-datameta-cv}

In Section \ref{comparing-competing-models}, following \citet[Secs. 5.1,
5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
the relationship of \code{mpg} to \code{horsepower} for the \code{Auto}
data, saving the 10 models, named \code{m.1} through \code{m.10}, in the
list \code{mlist}. We then used \code{cv()} to compare the
cross-validated MSE for the 10 models, discovering that the 7th degree
polynomial had the smallest MSE (by a small margin).

@@

Suppose that we want to cross-validate a multinomial logistic regression
model fit by the \code{multinom()} function in the \pkg{nnet} package
\citep{VenablesRipley:2002}. We borrow an example from \citet[Sec.
14.2.1]{Fox:2016}, with data from the British Election Panel Study on
vote choice in the 2001 UK election. Data for the example are in the
\code{BEPS} data set in the \pkg{carData} package:

@@

\begin{itemize}
\tightlist
\item
  \code{age}, in years.
\item
  \code{econ.cond.national} and \code{econ.cond.household}, the
  respondent's ratings of the state of the economy, on 1 to 5 scales.
\item
  \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders
  of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5
  scales.
\item
  \code{Europe}, an 11-point scale on attitude towards European
  integration, with high scores representing ``Euro-skepticism.''
\item
  \code{political.knowledge}, knowledge of the parties' positions on
  European integration, with scores from 0 to 3.
\item
  \code{gender}, \code{"female"} or \code{"male"}.
\end{itemize}

@@

Figure \ref{fig:BEPS-plot} shows an ``effect plot,'' using the the
\pkg{effects} package \citep{FoxWeisberg:2019} to visualize the
\code{Europe} \(\times\) \code{political.knowledge} interaction in a
``stacked-area'' graph:

@@

{\centering \includegraphics{Figures/BEPS-plot-1}

@@

As political knowledge increases, voters tend to align their votes more
closely with the party positions on European integration: The
Conservative Party was relatively Euro-skeptic, while Labour and the
Liberal Democrats were more supportive of the UK's participation in the
EU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function (y, yhat)
{
    if (!all(y %in% c(0, 1)))
        stop("response values not all 0 or 1")
    if (any(yhat < 0) || any(yhat > 1))
        stop("fitted values outside of interval [0, 1]")
    yhat <- round(yhat)
    result <- mean(y != yhat)
    attr(result, "casewise loss") <- "y != round(yhat)"
    result
}
<bytecode: 0x14fd342e8>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection{Other R software}\label{other-r-software}

@@

\subsection{Comparing cv to other software for cross-validation: SAS,
Stata, and
Python}\label{comparing-cv-to-other-software-for-cross-validation-sas-stata-and-python}

@@

\subsubsection{SAS}\label{sas}

@@

\subsubsection{Stata}\label{stata}

@@

\subsubsection{Python}\label{python}

@@

The weighted-least-squares (``WLS'') estimator of \(\boldsymbol{\beta}\)
is \citep[see, e.g.,][Sec. 12.2.2]{Fox:2016} \footnote{This is a
  definitional formula, which assumes that the model matrix
  \(\mathbf{X}\) is of full column rank, and which can be subject to
  numerical instability when \(\mathbf{X}\) is ill-conditioned.
  \code{lm()} uses the QR decomposition of the model matrix with
  pivoting to obtain computationally more stable results.} \[
\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1}
  \mathbf{X}^\top \mathbf{W} \mathbf{y}
\]

@@

\code{glm()} computes the maximum-likelihood estimates for a generalized
linear model by iterated weighted least squares \citep[see, e.g.,][Sec.
6.12]{FoxWeisberg:2019}. The last iteration is therefore just a WLS fit
of the ``working response'' on the model matrix using ``working
weights.'' Both the working weights and the working response at
convergence are available from the information in the object returned by
\code{glm()}.

>>>>

Validating research by replication on independently collected data is a
common scientific norm. Emulating this process in a single study by
data-division is less common \citep[see, e.g.,][]{Barnard:1974}: The
data are randomly divided into two, possibly equal-size, parts; the
first part is used to develop and fit (in the language of machine
leaning, ``train'') a statistical model; and then the second part is
used to assess the adequacy of (``validate'') the model fit to the first
part of the data. Data-division, however, suffers from two problems: (1)
Dividing the data decreases the sample size and thus increases sampling
error; and (2), even more disconcertingly, the results can vary
substantially based on the random division of the data, particularly in
smaller samples. See \citet[Sec.~5.3]{Harrell:2015} for cogent remarks
about data-division and cross-validation.

@@

The \code{cv()} function is introduced in the context of a preliminary
example in Section~\ref{preliminary-example-polynomial-regression} of
the paper.

\begin{itemize}
\item
  The \code{"lm"} and \code{"glm"} methods are for linear and
  generalized-linear models fit respectively by the standard
  \proglang{R} \code{lm()} and \code{glm()} functions.
\item
  The \code{"modList"} method for \code{cv()} cross-validates several
  competing models, not necessarily of the same class, using the same
  division of the data into folds.
\item
  Cross-validating mixed-effects models, using the \code{"glmmTMB"},
  \code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves
  special considerations that we take up in Section
  \ref{cross-validating-mixed-effects-models}.
\item
  The \code{"function"} method for \code{cv()}, discussed in Section
  \ref{cross-validating-model-specification}, cross-validates a
  potentially complex model-specification process that may, for example,
  involve choice of data transformations and predictors, or model
  selection via CV itself.
\item
  The \code{"default"} \code{cv()} method works (perhaps with a bit of
  coaxing) with many other existing regression-model classes for which
  there is an \code{update()} method that accepts a \code{data}
  argument. More generally, the \pkg{cv} package is designed to be
  extensible, as discussed in Section~\ref{extending-the-cv-package}.
\end{itemize}

@@

{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{jss5535_files/figure-latex/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{jss5535_files/figure-latex/mpg-horsepower-polynomials-2} }

@@

We'll focus on the relationship of \code{mpg} (miles per gallon) to
\code{horsepower}, as displayed in
Figure~\ref{fig:mpg-horsepower-polynomials} (a). The relationship
between the two variables is monotone, decreasing, and nonlinear.
Following \citet{JamesEtAl:2021}, we'll consider approximating the
relationship by a polynomial regression, with the degree of the
polynomial \(p\) ranging from 1 (a linear regression) to 10.\footnote{Although
  it serves to illustrate the use of CV, a polynomial is not the best
  choice here. Consider, for example, the scatterplot for
  log-transformed \code{mpg} and \code{horsepower}, produced by
  \code{plot(mpg $\sim$ horsepower, data = Auto, log = "xy")} (execution
  of which is left to the reader). We revisit the \code{Auto} data in
  Section~\ref{cross-validating-model-specification}.} Polynomial fits
for \(p = 1\) to \(5\) are shown in Figure
\ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly
inappropriate; the fits for \(p = 2\) (quadratic) through \(4\) are very
similar; and the degree-5 polynomial may over-fit the data by chasing
one or two relatively high \code{mpg} values at the right (but see the
CV results reported below).

Figure~\ref{fig:mpg-horsepower-polynomials} (b) shows two measures of
estimated (squared) error as a function of polynomial-regression degree:
The mean-squared error (``MSE''), defined as
\(\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\), and
the usual residual variance, defined as
\(\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2\).
Here \(y_i\) is the response and \(\widehat{y}_i\) the fitted value for
the \(i\)th case. The former necessarily declines with \(p\) (or, more
strictly, can't increase with \(p\)), while the latter gets slightly
larger for the largest values of \(p\), with the ``best'' value, by a
small margin, for \(p = 7\).

@@

{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{jss5535_files/figure-latex/polynomial-regression-CV-graph-2-1} }~~~\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{jss5535_files/figure-latex/polynomial-regression-CV-graph-2-2} }

@@

Following their use by \citet{RaudenbushBryk:2002}, data from the 1982
\emph{High School and Beyond} (``HSB'') survey have become a staple of
the literature on mixed-effects models. The HSB data are used by
\citet[Sec.~7.2.2]{FoxWeisberg:2019} to illustrate the application of
linear mixed models to hierarchical data, and we'll closely follow their
example here.

@@

In the process, we merge variables from the school-level and
student-level data sets, and create two new school-level variables:
\code{mean.ses}, which is the average SES for students in each school;
and \code{cses}, which is the individual students' SES centered at their
school means. For details, see \citet[Sec.~7.2.2]{FoxWeisberg:2019}.

@@

To generate the four data sets, we first constructed a template data set
from a population with a between-patient effect of dosage of \(1.0\) and
a within-patient effect of \(-0.5\). We then applied four different
multipliers to the between-patient standard deviation from the
regression line of patient centroids. The resulting data sets, shown in
Figure~\ref{fig:plot1}, differ in the between-patient variation of
patient centroids from a common between-patient regression line,
exhibiting a progression from low to high variation around the common
regression line. Each patient has the same relative configuration of
dosages and symptoms in each of the four data sets. To help visualize
the structure of the data, estimated 50\% Gaussian concentration
ellipses are shown for each patient.

@@

{\centering \includegraphics[width=0.75\linewidth]{jss5535_files/figure-latex/plot1-1}

@@

Using data like these, a researcher may attempt to obtain an estimate of
the within-patient effect of dosage by fitting mixed models with a
random intercept. We will illustrate how this approach results in
estimates that are highly sensitive to the relative variation at the two
levels of the mixed model by considering four models, all of which
include a random intercept. The first three models are sequentially
nested in their fixed effects: (1) \code{~ 1}, intercept only; (2)
\code{~ 1 + x}, intercept and effect of \code{x}; and (3)
\code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual
variable, \code{xm}, consisting of the within-patient mean of \code{x}.
The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the
centered-within-group variable \code{x - xm}. We thus fit four models to
each of the four data sets for a total of 16 models.

We proceed to obtain predictions from each model (1) using only the
fixed effects, as for cross-validation based on clusters (i.e.,
patients), and (2) using both fixed and random effects---that is, the
BLUPs---as for cross-validation based on cases (i.e., occasions within
patients). The fixed- and random-effects predictions from these models
are displayed in Figure~\ref{fig:plot-fits-combined} along with a
summary of the data using 50\% concentration ellipses (see the top row
of the figure).

@@

{\centering \includegraphics[width=1\linewidth]{jss5535_files/figure-latex/plot-fits-combined-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{jss5535_files/figure-latex/cross-validation-data-plot-1}

@@

In summary, when between-cluster variation is relatively large, the
model \code{~ 1 + x}, with \code{x} alone and without the contextual
mean \code{xm} of \code{x}, is assessed as fitting very poorly by
cluster-based CV, but relatively much better by case-based CV. In all of
our examples, the model \code{~ 1 + x + xm}, which includes both
\code{x} and its contextual mean, produces better results using both
cluster-based and case-based CV. These conclusions are consistent with
our observations based on graphing predictions from the various models
(in Figure~\ref{fig:plot-fits-combined} on page
\pageref{fig:plot-fits-combined}), and they illustrate the desirability
of assessing mixed-effect models at different hierarchical levels.

@@

As \citet[Sec.~7.10.2: ``The Wrong and Right Way to Do
Cross-validation'']{HastieTibshiraniFriedman:2009} explain, if the whole
data are used to specify or fine-tune a statistical model, then
subsequent cross-validation of the model is intrinsically misleading,
because the model is selected to fit the whole data, including the part
of the data that remains when each fold is removed. Statistical modeling
is partly a craft, and one could imagine applying that craft
independently to successive partial data sets, each with a fold removed.
The resulting procedure would be tedious, though possibly worth the
effort, but it would also be difficult to realize in practice: After
all, we can hardly erase our memory of statistical modeling choices
between analyzing partial data sets. Alternatively, if we're able to
automate the process of model specification, then we can more
realistically apply CV mechanically.

@@

\subsection[Example: Data transformation and predictor selection for the Auto
data]{Example: Data transformation and predictor selection for the
\code{Auto}
data}\label{example-data-transformation-and-predictor-selection-for-the-auto-data}

@@

The \code{Auto} data appeared in a preliminary example in Section
\ref{preliminary-example-polynomial-regression}, where we employed CV to
inform the selection of the degree of a polynomial regression of
\code{mpg} on \code{horsepower}. Here, we consider more generally the
problem of predicting \code{mpg} from the other variables in the
\code{Auto} data. We begin with a bit of data management, and then
examine the pairwise relationships among the numeric variables in the
data set (Figure~\ref{fig:Auto-explore-scatterplot-matrix}, produced by
the \code{scatterplotMatrix()} function in the \pkg{car} package); in
each panel, the solid line shows the linear least-squares fit and the
broken line is for a nonparametric regression:

@@

{\centering \includegraphics[width=0.95\linewidth]{jss5535_files/figure-latex/Auto-explore-scatterplot-matrix-1}

@@

{\centering \includegraphics[width=0.7\linewidth]{jss5535_files/figure-latex/Auto-explore-cr-plots-1}

@@

Some background: As \citet[Sec.~8.2]{Weisberg:2014} explains, there are
technical advantages to having (numeric) predictors in linear regression
analysis that are themselves linearly related. If the predictors
\emph{aren't} linearly related, then the relationships between them can
often be straightened by power transformations. Transformations can be
selected after graphical examination of the data, or by analytic
methods, such as transforming the predictors towards multivariate
normality, which implies linearity. Once the relationships between the
predictors are linearized, it can be advantageous similarly to transform
the conditional distribution of the response variable towards normality.
Selecting transformations analytically raises the possibility of
automating the process, as required for cross-validation.

The \code{powerTransform()} function in the \pkg{car} package transforms
variables towards multivariate normality by a generalization of Box and
Cox's maximum-likelihood-like approach \citep{BoxCox:1964}. Although the
multinormal likelihood is maximized, the variables are transformed
individually. Several ``families'' of power transformations can be used,
including the original Box-Cox family (which is the default), simple
powers (and roots), and two adaptations of the Box-Cox family to data
that may include negative values and zeros: the Box-Cox-with-negatives
family and the Yeo-Johnson family; see \citet[Chap.~8]{Weisberg:2014}
and \citet[Chap.~3]{FoxWeisberg:2019} for details. We proceed to
transform the numeric predictors in the \code{Auto} regression towards
multivariate normality:

@@

\subsection[Example: Applying meta CV to polynomial regression for the
Auto data]{\texorpdfstring{Example: Applying meta CV to polynomial
regression for the \code{Auto}
data\footnote{What we call here ``meta CV'' has also been termed
  ``nested CV'' \citep[e.g., in the \pkg{cvms} package
  by][]{OlsenZachariae:2024}. We prefer, however, to reserve the term
  ``nested CV'' for the procedure described by
  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying meta CV to polynomial regression for the Auto data}}\label{example-applying-meta-cv-to-polynomial-regression-for-the-auto-datameta-cv}

In Section~\ref{comparing-competing-models}, following \citet[Secs. 5.1,
5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
the relationship of \code{mpg} to \code{horsepower} for the \code{Auto}
data, saving the 10 models, named \code{m.1} through \code{m.10}, in the
object \code{mlist}. We then used \code{cv()} to compare the
cross-validated MSE for the 10 models, discovering that the 7th degree
polynomial had the smallest MSE (by a small margin).

@@

Suppose that we want to cross-validate a multinomial logistic regression
model fit by the \code{multinom()} function in the \pkg{nnet} package
\citep{VenablesRipley:2002}. We borrow an example from \citet[Sec.~14.2.1]{Fox:2016},
with data from the British Election Panel Study on
vote choice in the 2001 UK election. Data for the example are in the
\code{BEPS} data set in the \pkg{carData} package:

@@

\begin{itemize}
\tightlist
\item
  \code{age}, in years.
\item
  \code{econ.cond.national} and \code{econ.cond.household}, the
  respondent's ratings of the state of the economy, on 1 to 5 scales.
\item
  \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders
  of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5
  scales.
\item
  \code{Europe}, an 11-point scale for attitude towards European
  integration, with high scores representing ``Euro-skepticism.''
\item
  \code{political.knowledge}, knowledge of the parties' positions on
  European integration, with scores from 0 to 3.
\item
  \code{gender}, \code{"female"} or \code{"male"}.
\end{itemize}

@@

Figure~\ref{fig:BEPS-plot} shows an ``effect plot,'' using the the
\pkg{effects} package \citep{FoxWeisberg:2019} to visualize the
\code{Europe} \(\times\) \code{political.knowledge} interaction in a
``stacked-area'' graph:

@@

{\centering \includegraphics{jss5535_files/figure-latex/BEPS-plot-1}

@@

As political knowledge increased, voters tended to align their votes
more closely with the party positions on European integration: The
Conservative Party was relatively Euro-skeptic, while Labour and the
Liberal Democrats were more supportive of the UK's participation in the
EU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function (y, yhat)
{
    if (!all(y %in% c(0, 1)))
        stop("response values not all 0 or 1")
    if (any(yhat < 0) || any(yhat > 1))
        stop("fitted values outside of interval [0, 1]")
    yhat <- round(yhat)
    result <- mean(y != yhat)
    attr(result, "casewise loss") <- "y != round(yhat)"
    result
}
<bytecode: 0x11188dee8>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection[Other R software]{Other \proglang{R} software}\label{other-r-software}

@@

\subsection[Comparing cv to other software for cross-validation: SAS,
Stata, and Python]
{Comparing cv to other software for cross-validation: \proglang{SAS},
\proglang{Stata}, and \proglang{Python}}
\label{comparing-cv-to-other-software-for-cross-validation-sas-stata-and-python}

@@

\subsubsection[SAS]{\proglang{SAS}}\label{sas}

@@

\subsubsection[Stata]{\proglang{Stata}}\label{stata}

@@

\subsubsection[Python]{\proglang{Python}}\label{python}

@@

The weighted-least-squares (``WLS'') estimator of \(\boldsymbol{\beta}\)
is \citep[see, e.g.,][Sec.~12.2.2]{Fox:2016} \footnote{This is a
  definitional formula, which assumes that the model matrix
  \(\mathbf{X}\) is of full column rank, and which can be subject to
  numerical instability when \(\mathbf{X}\) is ill-conditioned.
  \code{lm()} uses the QR decomposition of the model matrix with
  pivoting to obtain computationally more stable results.} \[
\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1}
  \mathbf{X}^\top \mathbf{W} \mathbf{y}
\]

@@

\code{glm()} computes the maximum-likelihood estimates for a generalized
linear model by iterated weighted least squares \citep[see, e.g.,][Sec.~6.12]{FoxWeisberg:2019}.
The last iteration is therefore just a WLS fit
of the ``working response'' on the model matrix using ``working
weights.'' Both the working weights and the working response at
convergence are available from the information in the object returned by
\code{glm()}.
